---
title: 'Data Analysis Capstone Project'
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
---

Load the data and clean-up header column name 
```{r}
data <- read.csv("C:/Users/mfbro/Downloads/WA_Fn-UseC_-HR-Employee-Attrition.csv", header=TRUE, sep=",")
data$Age <- data$ï..Age
data <- data[-1]
```
Examine the structure 
```{r}
str(data)
```
Examine summary structure 
```{r}
summary(data)
```
Examine each variable using plot or hist
```{r}
hist(data$Age)
plot(data$Attrition)
plot(data$BusinessTravel)
hist(data$DailyRate)
plot(data$Department)
hist(data$DistanceFromHome)
hist(data$Education)
plot(data$EducationField)
plot(data$EmployeeCount)
plot(data$EmployeeNumber)
hist(data$EnvironmentSatisfaction)
plot(data$Gender)
hist(data$HourlyRate)
hist(data$JobInvolvement)
hist(data$JobLevel)
plot(data$JobRole)
hist(data$JobSatisfaction)
plot(data$MaritalStatus)
hist(data$MonthlyIncome)
hist(data$MonthlyRate)
hist(data$NumCompaniesWorked)
plot(data$Over18)
plot(data$OverTime)
hist(data$PercentSalaryHike)
hist(data$PerformanceRating)
hist(data$RelationshipSatisfaction)
plot(data$StandardHours)
hist(data$StockOptionLevel)
hist(data$TotalWorkingYears)
hist(data$TrainingTimesLastYear)
hist(data$WorkLifeBalance)
hist(data$YearsAtCompany)
hist(data$YearsInCurrentRole)
hist(data$YearsSinceLastPromotion)
hist(data$YearsWithCurrManager)
```
Findings from the graph analysis are as follows: 

Age - employee are young (mean age is 36 years) 

Attrition - attrition rate is 16% 

Business Travel - 80% of employees travel rarely or not at all

Daily rate - distribution is flat  

Department - 65% of employees work in R&D, 30% in Sales  

Distance from home - most employees work < 10km from home

Education - 65% of employees have bachelor degree or better 

Education Field - 80% of employees have a technical degree 

EmployeeCount - only one value, will be removed 

EmployeeNumber- internal number, will be removed 

Environment Satisfaction - most employees have a high environment satisfaction 

Gender - 60% men, 40% women

Hourly Rate - distribution is flat 

Job Involvement - most employees have a high job involvement 

Job level - the job level for most employees is fairly low 

Job Role - 17% management, 22% sales, technical 60%

Job Satisfaction - most employees have a high job satisfaction  

Maritial status - 22% divorced, 46% married, 31% single

Monthly Income - distribution is skewed, most employees have a low monthly income (median = 4900 vs mean = 6500)

Monthly rate - distribution is flat 

Num Companies Worked - most employees have only worked for a few companies (median = 2)

Over18 - only one value, will be removed 

Overtime - 30% of employees get overtime, 70% do not get overtime 

Percent salary Hike - most employees got the average salary 

Performance Rating - distribution is skewed, there are only 3 and 4 ratings 

Relationship Satisfaction - most employees are satisfied with relationship

StandardHours - only one value, will be removed

Stock Options Levels - 80% of employees get zero or few stock options 

Total Working Years - most employees have about 10 years work experience 

Training Times - most employees were trained 2 - 3 times in the year 

Work Life Balance - most employees are happy with work life balance 

Years at Company - 80% of employees have been with the company less than 10 years 

Years in Role - most employees have been in their role less than 5 years 

Years since promotion - most employees have been promoted within last years 

Years with manager - most employees have been with their manager less than 5 years 


```{r}

```

Check for missing data 
```{r}
sum(is.na(data))
```
Prepare Bar Charts against attrition against for each remaining variable  
```{r}
plot_age = table(data$Attrition, data$Age)
barplot(plot_age, main="Age vs Attrition", xlab="Age", legend=rownames(plot_age), beside = TRUE)
#
plot_travel = table(data$Attrition, data$BusinessTravel)
barplot(plot_travel, main = "Travel Frequency vs Attrition", xlab = "Travel Frequency", legend = rownames(plot_travel), beside = TRUE)
#
plot_dailyrate = table(data$Attrition, data$DailyRate)
barplot(plot_dailyrate, main = "Daily Rate vs Attrition", xlab = "Daily Rate", legend=rownames(plot_dailyrate), beside = TRUE)
#
plot_dept = table(data$Attrition, data$Department)
barplot(plot_dept, main = "Department vs Attrition", xlab ="Dept", legend=rownames(plot_dept), beside = TRUE)
#
plot_distance = table(data$Attrition, data$DistanceFromHome)
barplot(plot_distance, main = "Distance from Home vs Attrition", xlab="Distance from Home", legend=rownames(plot_distance), beside = TRUE)
#
plot_education = table(data$Attrition, data$Education)
barplot(plot_education, main ="Education vs Attrition", xlab="Education", legend=rownames(plot_education), beside = TRUE)
#
plot_field = table(data$Attrition, data$EducationField)
barplot(plot_field, main = "Education Field vs Attrition", xlab="Education Field", legend=rownames(plot_field), beside = TRUE)
#
plot_envsat = table(data$Attrition, data$EnvironmentSatisfaction)
barplot(plot_envsat, main="Environment Satisfaction vs Attrition", xlab="Environment Satisfaction", legend = rownames(plot_envsat), beside = TRUE)
#
plot_gender = table(data$Attrition, data$Gender)
barplot(plot_gender, main="Gender vs Attrition", xlab = "Gender", legend=rownames(plot_gender), beside=TRUE)
#
plot_hourlyrate = table(data$Attrition, data$HourlyRate)
barplot(plot_hourlyrate, main="Hourly rate vs Attrition", xlab="Hourly Rate", legend=rownames(plot_hourlyrate), beside = TRUE)
#
plot_involvement = table(data$Attrition, data$JobInvolvement)
barplot(plot_involvement, main="Job Involvement vs Attrition", xlab="Job Involvement", legend=rownames(plot_involvement), beside = TRUE)
#
plot_joblevel = table(data$Attrition, data$JobLevel)
barplot(plot_joblevel, main="Job Level", xlab="Job Level", legend=rownames(plot_joblevel), beside = TRUE)
#
plot_jobrole = table(data$Attrition, data$JobRole)
barplot(plot_jobrole, main="Job Role vs Attrition", xlab = "Job Role", legend=rownames(plot_jobrole), beside = TRUE)
#
plot_jobsat = table(data$Attrition, data$JobSatisfaction)
barplot(plot_jobsat, main="Job Satisfaction vs Attrition", xlab="Job Satisfaction", legend=rownames(plot_jobsat), beside = TRUE)
#
plot_maritial = table(data$Attrition, data$MaritalStatus)
barplot(plot_maritial, main="Maritial Status vs Attrition", xlab="Maritial Status", legend=rownames(plot_maritial), beside = TRUE)
#
plot_monthlyincome = table(data$Attrition, data$MonthlyIncome)
barplot(plot_monthlyincome, main="Monthly Income vs Attrition", xlab = "Monthly Income", legend=rownames(plot_monthlyincome), beside = TRUE)
#
plot_monthlyrate = table(data$Attrition, data$MonthlyRate)
barplot(plot_monthlyrate, main="Monthly Rate vs Attrition", xlab = "Monthly Income", legend=rownames(plot_monthlyincome), beside = TRUE)
#
plot_numcompany = table(data$Attrition, data$NumCompaniesWorked)
barplot(plot_numcompany, main="Number of Company Worked vs Attrition", xlab="Number of Company Worked", legend=rownames(plot_numcompany), beside = TRUE)
#
plot_overtime = table(data$Attrition, data$OverTime)
barplot(plot_overtime, main="Overtime vs Attrition", xlab="Overtime", legend=rownames(plot_overtime), beside = TRUE)
#
plot_salaryhike = table(data$Attrition, data$PercentSalaryHike)
barplot(plot_salaryhike, main="Percent Salary Hike vs Attrition", xlab="Percent Salary Hike", legend=rownames(plot_salaryhike), beside = TRUE)
#
plot_rating = table(data$Attrition, data$PerformanceRating)
barplot(plot_rating, main="Performance Rating vs Attrition", xlab="Performance Rating", legend=rownames(plot_rating), beside = TRUE)
#
plot_relsat = table(data$Attrition, data$RelationshipSatisfaction)
barplot(plot_relsat, main="Relationship Satisfaction vs Attrition", xlab="Relationship Satisfaction", legend=rownames(plot_relsat), beside = TRUE)
#
plot_options = table(data$Attrition, data$StockOptionLevel)
barplot(plot_options, main="Stock Option Level vs Attrition", xlab="Stock Options Level", legend=rownames(plot_options), beside = TRUE)
#
plot_totalworkyears = table(data$Attrition, data$TotalWorkingYears)
barplot(plot_totalworkyears, main ="Total Working Years vs Attrition", xlab = "Total Working Years ", legend=rownames(plot_totalworkyears), beside = TRUE)
#
plot_training = table(data$Attrition, data$TrainingTimesLastYear)
barplot(plot_training, main="Training Amount vs Attrition", xlab="Training Amount", legend=rownames(plot_training), beside = TRUE)
#
plot_worklife = table(data$Attrition, data$WorkLifeBalance)
barplot(plot_worklife, main="Worklife balance vs Attrition", xlab="Worklife balance", legend=rownames(plot_worklife), beside = TRUE)
#
plot_yearscompany = table(data$Attrition, data$YearsAtCompany)
barplot(plot_yearscompany, main="Years at Company vs Attrition", xlab="Years at Company", legend=rownames(plot_yearscompany), beside = TRUE)
#
plot_yearsrole = table(data$Attrition, data$YearsInCurrentRole)
barplot(plot_yearsrole, main="Year in Role vs Attrition", xlab="Years in Role", legend = rownames(plot_yearsrole), beside = TRUE)
#
plot_yearsmanager = table(data$Attrition, data$YearsWithCurrManager)
barplot(plot_yearsmanager, main="Years with manager vs Attrition", xlab = "Years with manager", legend=rownames(plot_yearsmanager), beside = TRUE)
#
plot_promotion = table(data$Attrition, data$YearsSinceLastPromotion)
barplot(plot_promotion, main="Years since promotion vs Attrition", xlab="Years since promotion", legend=rownames(plot_promotion), beside = TRUE)
```
Prepare frequency tables against attrition for each categorical variable
```{r}
plot_travel = table(data$Attrition, data$BusinessTravel)
prop.table(plot_travel, 2) 
#
plot_dept = table(data$Attrition, data$Department)
prop.table(plot_dept, 2) 
#
plot_education = table(data$Attrition, data$Education)
Names = c("Below College", "College", "bachelor", "Masters", "PHD")
colnames(plot_education) <- Names
prop.table(plot_education, 2)
#
plot_field = table(data$Attrition, data$EducationField)
prop.table(plot_field, 2)
#
plot_envsat = table(data$Attrition, data$EnvironmentSatisfaction)
Names = c("Low", "Medium", "High", "Very High")
colnames(plot_envsat) <- Names
prop.table(plot_envsat, 2)
#
plot_gender = table(data$Attrition, data$Gender)
prop.table(plot_gender, 2)
#
plot_involvement = table(data$Attrition, data$JobInvolvement)
Names = c("Low", "Medium", "High", "Very High")
colnames(plot_involvement) <- Names
prop.table(plot_involvement, 2)
#
plot_joblevel = table(data$Attrition, data$JobLevel)
Names = c("Lowest", "Low", "Medium", "High", "Highest")
colnames(plot_joblevel) <- Names
prop.table(plot_joblevel, 2)
#
plot_jobrole = table(data$Attrition, data$JobRole)
prop.table(plot_jobrole, 2)
#
plot_jobsat = table(data$Attrition, data$JobSatisfaction)
Names = c("Low", "Medium", "High", "Very High")
colnames(plot_jobsat) <- Names
prop.table(plot_jobsat, 2)
#
plot_maritial = table(data$Attrition, data$MaritalStatus)
prop.table(plot_maritial, 2)
#
plot_overtime = table(data$Attrition, data$OverTime)
prop.table(plot_overtime, 2)
#
plot_rating = table(data$Attrition, data$PerformanceRating)
Names = c("Excellent", "Outstanding")
colnames(plot_rating) <- Names
prop.table(plot_rating, 2)
#
plot_relsat = table(data$Attrition, data$RelationshipSatisfaction)
Names = c("Low", "Medium", "High", "Very High")
colnames(plot_relsat) <- Names
prop.table(plot_relsat, 2)
#
plot_options = table(data$Attrition, data$StockOptionLevel)
Names = c("None", "Low", "Medium", "High")
colnames(plot_options) <- Names
prop.table(plot_options, 2)
#
plot_worklife = table(data$Attrition, data$WorkLifeBalance)
Names = c("Low", "Medium", "High", "Very High")
colnames(plot_worklife) <- Names
prop.table(plot_worklife, 2)
#
```
Prepare boxplots for each numerical variable vs attrition
```{r}
plot(x=data$Attrition, y=data$Age, main ="Attrition vs Age")
plot(x=data$Attrition, y=data$DailyRate, main ="Attrition vs Daily Rate")
plot(x=data$Attrition, y=data$DistanceFromHome, main ="Attrition vs Distance from Home")
plot(x=data$Attrition, y=data$HourlyRate, main = "Attrition vs Hourly Rate")
plot(x=data$Attrition, y=data$MonthlyIncome, main="Attrition vs Monthly Income")
plot(x=data$Attrition, y=data$MonthlyRate, main="Attrition vs Monthly Rate")
plot(x=data$Attrition, y=data$NumCompaniesWorked, main="Attrition vs Number Companies Worked")
plot(x=data$Attrition, y=data$PercentSalaryHike, main="Attrition vs Percent Salary Hike")
plot(x=data$Attrition, y=data$TotalWorkingYears, main="Attrition vs Total Working Years")
plot(x=data$Attrition, y=data$TrainingTimesLastYear, main="Attrition vs Training Times")
plot(x=data$Attrition, y=data$YearsAtCompany, main="Attrition vs Years at Company")
plot(x=data$Attrition, y=data$YearsInCurrentRole, main="Attrition vs Years in Current Role")
plot(x=data$Attrition, y=data$YearsSinceLastPromotion, main="Attrition vs Years since last Promotion")
plot(x=data$Attrition, y=data$YearsWithCurrManager, main="Attrition vs Years with Current manager")
```

```{r}
```
```
Results of the attrition analysis are as follows: 

Age - young employees have a higher attrition rate   

Business Travel - employees who travel frequently have highe 

Daily rate - no significant difference 

Department - Sales and Human Resources have higher attrition rates than R&D

Distance from Home - employees who commute farther have higher attrition 

Education - employees with lower education (no college, college, etc.) are somewhat more likely to quit 

Education Field - there are differences by field, people in Sales and HR most likley to quit 

Employee Count - N/A (variable will be removed)   

Employee Number - N/A (variable will be removed)     

Environment Satisfaction - employees with low environment satisfaction much more likley to quit  

Gender - male and femal employees quit at similar rates 

Hourly rate - no significant difference   

Job Involvement - employees with low job involvement much more likly to quit  

Job Level - employees at low levels more likley to quit 

Job Role - employees in Sales and Human Resources more likly to quit

Job satisfaction - employees with low job satisfaction levels more likley to quit 

Maritial Status - single employees more liklely to quit 

Monthly Income - employees with lower incomes more likly to quit 

Monthly rate - no significant difference 

Number of companies worked - employyes who have worked for few companies more likley to leave 

Over 18 - N/A (variable will be removed)   

Overtime - employees who get overtime much more likley to quit 

Percent salary hike - no significant differenc between employees who stay or leave 

Performance rating - no significant difference between emploees who stay and leave.  

Relationship Satisfaction - employees with very low relationship satisfaction more likley to quit 

Standard hours - N/A (variable will be removed)   

Stock Options - employees with no options more likley to quit 

Total working years - employees with fewer working years more likley to quit 

Training times - people with very little training more likley to quit

Worklife balance - employees with low worklife balance much more likley to quit 

Years at company - employees with fewer years more likley to quit 

Years in current role - employyes most likley to leave in first few years in role 

Years since last promotion - no significant difference between employees who stay or leave 

Years with current manager - employees most likley to quit first year 
```{r}

```
Examine correlation between the remaing varibles. 
Remove dependent variable  plus four redundant variables identified above  
```{r}
# remove varibles
mydata <- data.frame(data)
mydata <- mydata[-1]
mydata <- mydata[-7]
mydata <- mydata[-7]
mydata <- mydata[-18]
mydata <- mydata[-22]
# convert factors to numeric
mydata$BusinessTravel <- as.numeric(data$BusinessTravel)
mydata$Department <- as.numeric(mydata$Department)
mydata$EducationField <- as.numeric(mydata$EducationField)
mydata$Gender <- as.numeric(mydata$Gender)
mydata$JobRole <- as.numeric(mydata$JobRole)
mydata$MaritalStatus <- as.numeric(mydata$MaritalStatus)
mydata$OverTime <- as.numeric(mydata$OverTime)
# calculate correlation 
mydata.cor = cor(mydata)
round(mydata.cor,2)
```
Plot Correlations 
```{r}
install.packages("corrplot")
library(corrplot)
corrplot(mydata.cor)
```
An analysis of corr and corrplot shows that a number of variables seem to have strong (corr 0.60 - 0.79) or very strong (corr > 0.80) correlations:
- job role vs dept (corr = 0.66)
- job level vs monthly income (corr = 0.95)
- job level vs total working years (corr = 0.78)
- monthly income vs total working years (corr = 0.77)
- age vs total working years(corr = 0.68)
- percent salary hike vs performance rating (corr = 0.77)
- years at company vs total working years (corr =0.63)
- years at company vs years in current role (corr = 0.76)
- years with current manager vs years at company (corr =0.77)
- years with current manager vs years in current role (corr=0.77)
- years at company vs years since last promotion (corr =0.62)

```{r}

```

#
```{r}

```
# 

```{r}

```
#

```{r}

```
#
```{r}

```
#
```{r}

```
# 

